---
title: "Take-Home Exercise 1"
data: "2 December 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

## **Setting the Scene**

As urban systems digitize, encompassing buses, taxis, transit, utilities, and roads, the resulting datasets offer a comprehensive framework for tracing movement over both space and time. This surge in pervasive technologies like GPS and RFID integrated into vehicles amplifies this capability. For instance, smart cards and GPS devices in public buses collect extensive data on routes and ridership, potentially unveiling intricate patterns that divulge vital insights into measured phenomena. Analyzing and comparing these patterns could yield profound understanding of human behavior and movement within cities. These insights hold promise for enhancing urban management and providing crucial information to both public and private urban transport services for informed decision-making and competitive edge.

Despite the potential, the real-world application of this abundant location-aware data tends to be limited to basic tracking and mapping using GIS applications. This limitation stems from conventional GIS tools lacking the necessary functionality to effectively analyze and model spatial and spatio-temporal data.

## **Objectives**

Exploratory Spatial Data Analysis (ESDA) hold tremendous potential to address complex problems facing society. Applying Geovisualisation and Analysis and Local Indicators of Spatial Association (GLISA) to undercover the spatial and spatio-temporal mobility patterns of public bus passengers in Singapore.

## Getting started

Loading all the package that could be use for this into R.

```{r}
pacman::p_load(tmap, sf, sfdep, dplyr, mapview, ggpubr, tidyverse)
```

### **Geospatial Data**

Import the BusStop data, this data included the information of all the bus stops located in Singapore. This data is provided by LTA DATA MALL.

```{r}
BusStop <- st_read(dsn = "data/geospatial",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
tmap_mode("view")
```

Plotting the raw data, this plot show all the BusStops location

```{r}
tm_shape(BusStop) + 
  tm_dots()
```

### **Hexagonal Grid Creation**

```{r}
grid = st_make_grid(BusStop, c(500), what = "polygons", square = FALSE)
# sf and add grid ID
grid_sf = st_sf(grid) %>%
  # add grid ID
  mutate(grid_id = 1:length(lengths(grid)))
```

```{r}
grid_sf$n_colli = lengths(st_intersects(grid_sf, BusStop))

# remove grid without value of 0 (i.e. no points in side that grid)
grid_count = filter(grid_sf,n_colli > 0 )
```

```{r}
tmap_mode("view")
```

Plotting hexagon map

```{r}
tm_shape(grid_count) +
  tm_fill(
    col = "n_colli",  
    palette = "Blues",
    style = "cont",
    title = "Number of collisions",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
    popup.vars = c(
      "Number of collisions: " = "n_colli"
    ),
    popup.format = list(
      n_colli = list(format = "f", digits = 0)
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.7)
```

Next, as we can see there are some busstop that is outside Singapore. In this case it will be exclude.

Removing the Bus Stops outside Singapore

```{r}
grid_count_rm <- grid_count %>%
  filter(!grid_id == 1767,
         !grid_id == 2073,
         !grid_id == 2135,
         !grid_id == 2104)
```

New Plot after removed the Bus Stops outside Singapore

```{r}
tm_shape(grid_count_rm) +
  tm_fill(
    col = "n_colli",  
    palette = "Blues",
    style = "cont",
    title = "Number of collisions",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
    popup.vars = c(
      "Number of collisions: " = "n_colli"
    ),
    popup.format = list(
      n_colli = list(format = "f", digits = 0)
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.7)
```

### **Aspatial Data**

Import the aspatial data into R, the data that will be use here is the "origin_destination_bus_202310". This data is originally from the LTA DATA MALL.

```{r}
busTrips <- read_csv("data/aspatial/origin_destination_bus_202310.csv")
```

```{r}
# busTrips <- read_csv("data/aspatial/origin_destination_bus_202308.csv")
# busTrips <- read_csv("data/aspatial/origin_destination_bus_202309.csv")

busTrips$ORIGIN_PT_CODE <- as.factor(busTrips$ORIGIN_PT_CODE)
busTrips$DESTINATION_PT_CODE <- as.factor(busTrips$DESTINATION_PT_CODE)
```

## Data Wrangling 

### **Aspatial Data Wrangling**

Calculating bus trip in weekday and morning peak hour

```{r}
busTripsDayMorning <- busTrips %>%
  filter(DAY_TYPE == "WEEKDAY", 
         TIME_PER_HOUR >= 6, 
         TIME_PER_HOUR <= 9) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(WeekdayMorningTrips = sum(TOTAL_TRIPS))
```

Calculating bus trip in weekday and afternoon peak hour

```{r}
busTripsDayAfternoon <- busTrips %>%
  filter(DAY_TYPE == "WEEKDAY", 
         TIME_PER_HOUR >= 17, 
         TIME_PER_HOUR <= 20) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(WeekdayAfternoonTrips = sum(TOTAL_TRIPS))
```

Calculating bus trip in weekend and morning peak hour

```{r}
busTripsEndMorning <- busTrips %>%
  filter(DAY_TYPE == "WEEKENDS/HOLIDAY", 
         TIME_PER_HOUR >= 11, 
         TIME_PER_HOUR <= 14) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(WeekendMorningTrips = sum(TOTAL_TRIPS))
```

Calculating bus trip in weekend and evening peak hour

```{r}
busTripsEndEvening <- busTrips %>%
  filter(DAY_TYPE == "WEEKENDS/HOLIDAY", 
         TIME_PER_HOUR >= 16, 
         TIME_PER_HOUR <= 19) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(WeekendEveningTrips = sum(TOTAL_TRIPS))
```

Joining all the Peak Trips

```{r}
BusTrips_comb <- busTripsDayMorning %>%
  left_join(busTripsDayAfternoon) %>%
  left_join(busTripsEndMorning) %>%
  left_join(busTripsEndEvening)
```

### **Geospatial Data Wrangling**

```{r}
grid_bus <- st_join(grid_count_rm,BusStop,join = st_contains) %>%
  unnest() %>%
  select(grid_id,BUS_STOP_N)
```

```{r}
grid_bus$BUS_STOP_N <- as.factor(grid_bus$BUS_STOP_N)
```

### **Joining aspatial and geospatial**

Joining both data together

```{r}
Trips <- left_join(BusTrips_comb,grid_bus,
                   by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
  group_by(grid_id)%>%
  summarise(WeekdayMorningTrips = sum(WeekdayMorningTrips),
            WeekdayAfternoonTrips = sum(WeekdayAfternoonTrips),
            WeekendMorningTrips = sum(WeekendMorningTrips),
            WeekendEveningTrips = sum(WeekendEveningTrips))
```

```{r}
Trips <- left_join(grid_count_rm,Trips) %>%
  mutate (Total_Trips = WeekdayMorningTrips+WeekdayAfternoonTrips+WeekendMorningTrips  +WeekendEveningTrips) %>% 
  rename (n_bus = n_colli)
```

Trip amount per bus stop, to determine whether the quantity of bus stations or whether it is indeed crowded.

```{r}
TripsPerBusStop <- Trips %>%
  mutate (WeekdayMorningTrips = WeekdayMorningTrips/n_bus,
          WeekdayAfternoonTrips = WeekdayAfternoonTrips/n_bus,
          WeekendMorningTrips = WeekendMorningTrips/n_bus,
          WeekendEveningTrips = WeekendEveningTrips/n_bus)
```

```{r}
TripsLog <- Trips %>%
  mutate (WeekdayMorningTrips = log(WeekdayMorningTrips),
          WeekdayAfternoonTrips = log(WeekdayAfternoonTrips),
          WeekendMorningTrips = log(WeekendMorningTrips),
          WeekendEveningTrips = log(WeekendEveningTrips))
```

## **Visualising**

```{r}
tmap_mode("view")
```

The map below displays the total bus trip for each bus station of origin.

```{r}
tm_shape(Trips) +
  tm_fill(
    col = "Total_Trips",
    palette = "Blues",
    style = "quantile",
    title = "Total Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
    popup.format = list(
      Total_Trips = list(format = "f", digits = 0)
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.7)
```

It is evident from the above plot of total trips that bus travel is dispersed throughout the nation. On the other hand, many clusters are visible.

### **Total Trips**

```{r}
weekday_morning <- tm_shape(Trips) +
  tm_fill(
    col = "WeekdayMorningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekday Morning Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekday_afternoon <- tm_shape(Trips) +
  tm_fill(
    col = "WeekdayAfternoonTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekday Afternoon Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekend_morning <- tm_shape(Trips) +
  tm_fill(
    col = "WeekendMorningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekend Morning Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekend_evening <- tm_shape(Trips) +
  tm_fill(
    col = "WeekendEveningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekend Evening Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
tmap_mode("plot")
```

```{r}
tmap_arrange(weekday_morning, weekday_afternoon, weekend_morning, weekend_evening, ncol=2, nrow=2)
```

The segmented trips in the map above are broken down into four sector: weekend morning, weekend afternoon, and weekend morning. The distinctions between them are negligible. On the east side, there is a noticeable difference, though. On weekends and weekdays, the east side quantile is lower in the evening or afternoon than it is in the morning. Additionally, it is evident in the central that in the afternoon and evening, their quantile is higher.

### **Total Trips per Bus Stop**

```{r}
weekday_morning_per_stop <- tm_shape(TripsPerBusStop) +
  tm_fill(
    col = "WeekdayMorningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekday Morning Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekday_afternoon_per_stop <- tm_shape(TripsPerBusStop) +
  tm_fill(
    col = "WeekdayAfternoonTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekday Afternoon Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekend_morning_per_stop <- tm_shape(TripsPerBusStop) +
  tm_fill(
    col = "WeekendMorningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekend Morning Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekend_evening_per_stop <- tm_shape(TripsPerBusStop) +
  tm_fill(
    col = "WeekendEveningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekend Evening Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
tmap_mode("plot")
```

```{r}
tmap_arrange(weekday_morning_per_stop, weekday_afternoon_per_stop,
             weekend_morning_per_stop, weekend_evening_per_stop,
            ncol=2, nrow=2) 
```

The diagram above demonstrates how the overall number of trips varies depending on the day and time of day. Each bus station in a hexagon has a different amount of total trips within it. Still, the outcomes are strikingly identical to the last one.

### **Log value of trips**

```{r}
weekday_morning_log <- tm_shape(TripsLog) +
  tm_fill(
    col = "WeekdayMorningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekday Morning Trips per Bus Stop",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekday_afternoon_log <- tm_shape(TripsLog) +
  tm_fill(
    col = "WeekdayAfternoonTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekday Afternoon Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekend_morning_log <- tm_shape(TripsLog) +
  tm_fill(
    col = "WeekendMorningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekend Morning Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
weekend_evening_log <- tm_shape(TripsLog) +
  tm_fill(
    col = "WeekendEveningTrips",
    palette = "Blues",
    style = "quantile",
    title = "Weekend Evening Trips",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6
  ) +
  tm_borders(col = "grey40", lwd = 0.7) +
  tm_legend(scale = 0.5)
```

```{r}
tmap_mode("plot")
```

```{r}
tmap_arrange(weekday_morning_log, weekday_afternoon_log,
             weekend_morning_log, weekend_evening_log, ncol=2, nrow=2)
```

When the number is logarithmic, the value of the total number of trips for each hexagon is displayed in the figure above. Additionally, there aren't many differences between them.

## **Local Indicators of Spatial Association (LISA) Analysis**

### **Spatial Weight**

```{r}
wm_idw <- Trips %>%
  mutate(nb = st_dist_band(grid),
         wts = st_inverse_distance(nb, grid,
                                   scale = 1,
                                   alpha = 1),
         .before = 1) %>%
  mutate(grid_id = 1:length(Trips$grid_id))
```

### **Local Moran**

#### Weekday Morning

```{r}
lisa <- wm_idw %>% 
  mutate(local_moran = local_moran(
    WeekdayMorningTrips, nb, wts, nsim = 99, na.action=na.pass),
         .before = 1) %>%
  unnest(local_moran)
```

```{r}
ii_val_moran <- tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of Trip",
            main.title.size = 0.8)
```

```{r}
p_val_moran <- tm_shape(lisa) +
  tm_fill("p_ii_sim") + 
  tm_borders(alpha = 0.5) +
   tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)
```

```{r}
tmap_arrange(ii_val_moran, p_val_moran, ncol = 2)
```

```{r}
lisa_sig <- lisa  %>%
  filter(p_ii_sim < 0.05)
pal <- RColorBrewer::brewer.pal("Set1", n = 5)

tm_shape(lisa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(lisa_sig) +
  tm_fill("mean", palette = pal ) + 
  tm_borders(alpha = 0.4)
```

#### Weekday Afternoon

```{r}
lisa <- wm_idw %>% 
  mutate(local_moran = local_moran(
    WeekdayAfternoonTrips, nb, wts, nsim = 99, na.action=na.pass),
         .before = 1) %>%
  unnest(local_moran)
```

```{r}
ii_val_moran <- tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of Trip",
            main.title.size = 0.8)
```

```{r}
p_val_moran <- tm_shape(lisa) +
  tm_fill("p_ii_sim") + 
  tm_borders(alpha = 0.5) +
   tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)
```

```{r}
tmap_arrange(ii_val_moran, p_val_moran, ncol = 2)
```

```{r}
lisa_sig <- lisa  %>%
  filter(p_ii_sim < 0.05)
pal <- RColorBrewer::brewer.pal("Set1", n = 5)

tm_shape(lisa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(lisa_sig) +
  tm_fill("mean", palette = pal ) + 
  tm_borders(alpha = 0.4)
```

#### Weekend Morning 

```{r}
lisa <- wm_idw %>% 
  mutate(local_moran = local_moran(
    WeekendMorningTrips, nb, wts, nsim = 99, na.action=na.pass),
         .before = 1) %>%
  unnest(local_moran)
```

```{r}
ii_val_moran <- tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of Trip",
            main.title.size = 0.8)
```

```{r}
p_val_moran <- tm_shape(lisa) +
  tm_fill("p_ii_sim") + 
  tm_borders(alpha = 0.5) +
   tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)
```

```{r}
tmap_arrange(ii_val_moran, p_val_moran, ncol = 2)
```

```{r}
lisa_sig <- lisa  %>%
  filter(p_ii_sim < 0.05)
pal <- RColorBrewer::brewer.pal("Set1", n = 5)

tm_shape(lisa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(lisa_sig) +
  tm_fill("mean", palette = pal ) + 
  tm_borders(alpha = 0.4)
```

#### Weekend Evening

```{r}
lisa <- wm_idw %>% 
  mutate(local_moran = local_moran(
    WeekendEveningTrips, nb, wts, nsim = 99, na.action=na.pass),
         .before = 1) %>%
  unnest(local_moran)
```

```{r}
ii_val_moran <- tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of Trip",
            main.title.size = 0.8)
```

```{r}
p_val_moran <- tm_shape(lisa) +
  tm_fill("p_ii_sim") + 
  tm_borders(alpha = 0.5) +
   tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)
```

```{r}
tmap_arrange(ii_val_moran, p_val_moran, ncol = 2)
```

```{r}
lisa_sig <- lisa  %>%
  filter(p_ii_sim < 0.05)
pal <- RColorBrewer::brewer.pal("Set1", n = 5)

tm_shape(lisa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(lisa_sig) +
  tm_fill("mean", palette = pal ) + 
  tm_borders(alpha = 0.4)
```
