---
title: "In-Class Exercise 2: Spatial Weights, GLSA, EHSA"
data: "25 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

# In-Class Exercise 2: Spatial Weights, GLSA, EHSA

## Getting started 

Import R packages

```{r}
pacman::p_load(tmap, sf, sfdep, tidyverse, knitr, plotly)
```

## The Data 

Import data

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")

hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

```{r}
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
```

## Combining both data frame by using left join 

```{r}
hunan_GDPPC <- left_join(hunan,hunan2012)%>%
  select(1:4, 7, 15)
```

## Plotting a choropieth map

```{r}
tmap_mode("plot")
tm_shape(hunan_GDPPC) +
  tm_fill("GDPPC", 
          style = "quantile", 
          palette = "Blues",
          title = "GDPPC") +
  tm_layout(main.title = "Distribution of GDP per capita by district, Hunan Province",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2)
```

## **Deriving Contiguity Spatial Weights**

### **Identifying contiguity neighbours: Queen\'s method**

```{r}
nb_queen <- hunan_GDPPC %>% 
  mutate(nb = st_contiguity(geometry),
         .before = 1)
```

```{r}
summary(nb_queen$nb)
```

```{r}
kable(head(nb_queen,
           n=10))
```

### **Identify contiguity neighbours: Rooks\' method**

```{r}
nb_rook <- hunan_GDPPC %>% 
  mutate(nb = st_contiguity(geometry,
                            queen = FALSE),
         .before = 1)
```

```{r}
nb2_queen <-  hunan_GDPPC %>% 
  mutate(nb = st_contiguity(geometry),
         nb2 = st_nb_lag_cumul(nb, 2),
         .before = 1)
```

```{r}
nb2_queen
```

## **Deriving contiguity weights: Queen\'s method**

### **Identifying contiguity neighbours: Queen\'s method**

```{r}
nb_queen <- hunan_GDPPC %>% 
  mutate(nb = st_contiguity(geometry),
         .before = 1)
```

```{r}
summary(nb_queen$nb)
```

```{r}
nb_queen
```

```{r}
kable(head(nb_queen, n=10))
```

### **Identify contiguity neighbours: Rooks\' method**

```{r}
nb_rook <- hunan_GDPPC %>% 
  mutate(nb = st_contiguity(geometry,
                            queen = FALSE),
         .before = 1)
```

### **Identifying higher order neighbors**

```{r}
nb2_queen <-  hunan_GDPPC %>% 
  mutate(nb = st_contiguity(geometry),
         nb2 = st_nb_lag_cumul(nb, 2),
         .before = 1)
```

```{r}
nb2_queen
```

## **Deriving contiguity weights: Queen\'s method**

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry), 
         wt = st_weights(nb, style = "W"), .before = 1)
```

```{r}
wm_q
```

### **Deriving fixed distance weights**

```{r}
geo <- sf::st_geometry(hunan_GDPPC)
nb <- st_knn(geo, longlat = TRUE)
dists <- unlist(st_nb_dists(geo, nb))
```

```{r}
summary(dists)
```

```{r}
wm_fd <- hunan_GDPPC %>%
  mutate(nb = st_dist_band(geometry,
                           upper = 66),
               wt = st_weights(nb),
               .before = 1)
```

### **Deriving adaptive distance weights**

```{r}
wm_ad <- hunan_GDPPC %>% 
  mutate(nb = st_knn(geometry,
                     k=8),
         wt = st_weights(nb),
               .before = 1)
```

### **Calculate inverse distance weights**

```{r}
wm_idw <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wts = st_inverse_distance(nb, geometry,
                                   scale = 1,
                                   alpha = 1),
         .before = 1)
```

## **Computing Global Moran\' I**

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```

### **Performing Global Moran\'sI test**

```{r}
global_moran_test(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
```

### **Performing Global Moran\'I permutation t**

```{r}
set.seed(1234)
```

```{r}
global_moran_perm(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt,
                  nsim = 99)
```

## **Computing local Moran\'s I**

```{r}
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(local_moran)
```

### **Visualising local Moran\'s I**

```{r}
tmap_mode("plot")
tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of GDPPC",
            main.title.size = 0.8)
```

### **Visualising p-value of local Moran\'s I**

```{r}
tmap_mode("plot")
tm_shape(lisa) +
  tm_fill("p_ii_sim") + 
  tm_borders(alpha = 0.5) +
   tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)
```

### **Visuaising local Moran\'s I and p-value**

```{r}
tmap_mode("plot")
map1 <- tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of GDPPC",
            main.title.size = 0.8)

map2 <- tm_shape(lisa) +
  tm_fill("p_ii",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
              labels = c("0.001", "0.01", "0.05", "Not sig")) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)

tmap_arrange(map1, map2, ncol = 2)
```

### **Visualising LISA map**

```{r}
lisa_sig <- lisa  %>%
  filter(p_ii < 0.05)
tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(lisa_sig) +
  tm_fill("mean") + 
  tm_borders(alpha = 0.4)
```

## **Hot Spot and Cold Spot Area Analysis (HCSA)**

## **Computing local Gi\* statistics**

```{r}
wm_idw <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wts = st_inverse_distance(nb, geometry,
                                   scale = 1,
                                   alpha = 1),
         .before = 1)
```

```{r}
HCSA <- wm_idw %>% 
  mutate(local_Gi = local_gstar_perm(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_Gi)
HCSA
```

### **Visualising Gi\***

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8))
```

### **Visualising p-value of HCSA**

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("p_sim") + 
  tm_borders(alpha = 0.5)
```

### **Visuaising local HCSA**

```{r}
tmap_mode("plot")
map1 <- tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "Gi* of GDPPC",
            main.title.size = 0.8)

map2 <- tm_shape(HCSA) +
  tm_fill("p_value",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
          labels = c("0.001", "0.01", "0.05", "Not sig")) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of Gi*",
            main.title.size = 0.8)

tmap_arrange(map1, map2, ncol = 2)
```

## **Visualising hot spot and cold spot areas**

```{r}
HCSA_sig <- HCSA  %>%
  filter(p_sim < 0.05)
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(HCSA_sig) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.4)
```

# **Emerging Hot Spot Analysis**

## Creating a Time Series Cube 

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

```{r}
is_spacetime_cube(GDPPC_st)
```

## Deriving the spatial weights 

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_inverse_distance(nb, geometry,
                                  scale = 1,
                                  alpha = 1),
         .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")

```

```{r}
head(GDPPC_nb)
```

## **Computing Gi\***

```{r}
gi_stars <- GDPPC_nb %>%
  group_by(Year) %>%
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>%
  tidyr::unnest(gi_star)
```

# **Mann-Kendall Test**

```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

```{r}
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

```{r}
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
```

## **Arrange to show significant emerging hot/cold spots**

```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:5)
```

## **Performing Emerging Hotspot Analysis**

```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

### **Visualising the distribution of EHSA classes**

```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

### **Visualising EHSA**

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

```{r}
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(alpha = 0.4)
```
