---
title: "In-class_Exercise1"
data: "18 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

## Getting Started

The code chunk below load the following package:

-   tmap: for thematic mapping

-   sf: for geospatial data handling

-   tidyverse: for non-spatial data handling

```{r}
pacman::p_load(tmap, sf, tidyverse)
```

## Preparing the FLow Data

### Importing the OD data

Firstly, we will import the *Passenger Volume by Origin Destination Bus Stops* data set downloaded from LTA DataMall by using 'read_csv()' of **read** package.

```{r}
# eval: false 
odbus <- read_csv("data/aspatial/origin_destination_bus_202308.csv")
```

A quick check of odbus tibble data frame shows that the values

```{r}
# eval: false
odbus$ORIGIN_PT_CODE <-
as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <-
as.factor(odbus$DESTINATION_PT_CODE) 
```

```{r}
# eval: false 
origin_7_9 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 7 &
         TIME_PER_HOUR <= 9) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS))
```

Two geospatial data will be used in this exerise

```{r}
# eval: false 
busstop <- st_read(dsn = "data/geospatial", layer = "BusStop") %>% st_transform(crs = 3414)
```

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MPSZ-2019") %>% 
  st_transform(crs = 3414)
  
```
